---
title: "Obtain protein annotations"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: pdf_document
---
Here, we obtain uniprot details and GO annotations for the proteins quantified. To save this needing to be re-run if there are changes in how the output is processed to protein-level abundances, we will obtain the feature data for all proteins detected in the evidence files.

```{r, warning=FALSE, message=FALSE, echo=FALSE}

library(here)
library(dplyr)
library(tidyr)
library(biomasslmb)
library(QFeatures)
library(uniprotREST)
library(ggplot2)

```


  
  
```{r, warning=FALSE, message=FALSE}
maxQuant_outfiles <- Sys.glob(here('raw/*/evidence.txt*'))

```

```{r, warning=FALSE, message=FALSE}
all_proteins <- maxQuant_outfiles %>% lapply(function(x){
  read.delim(x) %>%
  pull(Leading.razor.protein) %>% 
  unique() %>%
  strsplit(';') %>%
  unlist() %>% unique()
}) %>% unlist() %>% unique()

length(all_proteins)
```

## Uniprot protein details

Below, we obtain the protein details from Uniprot using the uniprotREST package.

```{r, warning=FALSE, message=FALSE}
uniprot2details <- uniprot_map(
    all_proteins,
    from='UniProtKB_AC-ID',
    method='stream') %>%
  dplyr::rename('UniprotID'=From) %>%
  select(-Entry) %>%
  mutate(Gene.Names.First=gsub(' .*','', Gene.Names))

# Some UniprotIDs generate multiple rows. It looks like this is because they have been subsequently 'demerged'. Below, we merge them
uniprot2details_unique <- uniprot2details %>%
  group_by(UniprotID) %>%
  summarise_all(.funs=function(x) paste(x, collapse=';'))

saveRDS(uniprot2details_unique, here('external/uniprot2details_unique.rds'))
```



## GO terms

Below we obtain the GO terms from Uniprot using the API and then we reformat the output before expanding the GO terms to include all parent terms too.

```{r, warning=FALSE, message=FALSE}

tempfilename <- tempfile()

uniprot_map(
  ids = all_proteins,
  from = "UniProtKB_AC-ID",
  method='stream',
  to = "UniProtKB", compressed = TRUE,
  format = "tsv", path = tempfilename,
  fields = "go",
) 

go_res <- read.delim(tempfilename) %>% dplyr::rename(c('UNIPROTKB'='From'))

file.remove(tempfilename)

# The Uniprot API returns a data.frame with all the GO terms for each protein in a single column
# Below, we separate each GO terms to be on a different row
go_res <- go_res %>%
  filter(Gene.Ontology..GO.!='') %>% separate_rows(Gene.Ontology..GO., sep = '; ') %>%
  separate_wider_delim(Gene.Ontology..GO., ' [GO:', names=c('GO_desc', 'GO.ID')) %>%
  mutate(GO.ID=paste0('GO:', gsub(']$', '', GO.ID)))

# And then expand the GO terms so that all ancestor/parent terms are annotated to each protein
# This is to avoid issues with GO over-representation testing packages only using 
# the directly annotated terms and not the parent terms
go_res_all <- get_ancestor_go(go_res)
```




Save to file for use in other notebooks.

```{r}
saveRDS(go_res_all, here('external/go_res_all.rds'))

```


