---
title: "PER2-HALO: QC peptide-level quantification and filter"
author: "Tom Smith"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Purpose

This notebook reads the peptide-level output from the MaxQuant search of the PER2-HALO immunoprecipitation (IP) experiment (PER2-HALO fusion vs free HALO tag, 3 replicates each), applies routine peptide-level quality filtering, and normalises the abundances, producing a `QFeatures` object ready for protein-level summarisation and statistical testing in notebook 2.

The IP design means the two conditions are expected to differ substantially in their overall protein composition (a free HALO tag pulls down a different, and likely smaller, set of interactors than the PER2-HALO fusion), so particular attention is paid to the choice of normalisation strategy: standard median normalisation, which assumes the two conditions have a similar overall abundance distribution, is compared against normalisation to a set of proteins (histones) expected to be stable regardless of condition.

**Input:**

- `raw/HALO/modificationSpecificPeptides.txt.gz` — MaxQuant peptide-level (LFQ intensity) quantification.
- `raw/HALO/peptides.txt.gz` — used only for its `Sequence` to `Leading.razor.protein` mapping, to standardise protein assignment.
- `raw/HALO/proteinGroups.txt.gz` — protein-level search results, used here for the `Q.value` (FDR) column.
- `raw/HALO/mqpar.xml` — the MaxQuant parameter file, used to recover the mapping from MaxQuant's internal experiment labels to sample names.
- `external/uniprot2details_unique.rds` from notebook 0 — UniProt protein annotations.

**Output:**

- `notebooks/output/HALO_lfq_qf.rds` — a `QFeatures` object retaining every processing stage (raw, filtered, normalised, histone-normalised, and missing-value-filtered peptide assays), consumed by notebook 2.

**Analysis:**

1. **Read data** — read the peptide-level quantification, standardise the `Leading.razor.protein` assignment for peptides matching the transgene sequences (see below), and join in the protein annotations and FDR (`Q.value`) from the protein-level search.
2. **Sample mapping** — recover the mapping from MaxQuant's internal experiment names to condition/replicate from the search parameter file, and build the `QFeatures` object.
3. **Zero handling** — replace MaxQuant's reported zero intensities with `NA`, since mass spectrometry cannot assert a true zero abundance.
4. **Filtering** — remove contaminant peptides and peptides belonging to proteins that do not pass the 1% FDR threshold.
5. **Normalisation** — log2-transform, then compare standard median normalisation against normalisation to a set of stably-expressed histone proteins, and carry forward the histone-normalised assay as the one used downstream.
6. **Missing value inspection** — visualise missingness patterns and construct a peptide set filtered to a maximum of 4/6 missing values, primarily as a diagnostic of how much data a naive missing-value filter would discard.
7. **QC summary** — plot the number of peptides and proteins retained at each processing stage.
8. **Export** — save the full `QFeatures` object.

**How to read the results:** the assay actually carried forward into notebook 2 is `peptides_norm_hist` (histone-normalised, but *not* filtered for missing values). Notebook 2 summarises to protein level using `limpa`, which models missingness explicitly via a detection probability curve rather than requiring complete peptides, so there is no need to discard peptides with missing values before that step; the 4/6-missing-value-filtered assay (`peptides_filtered_missing`) constructed here is used only for the peptide/protein-count QC plots at the end of this notebook, not as an input to notebook 2.

# Setup

Load required packages

```{r, message=FALSE}
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
library(here)
library(QFeatures)
library(xml2)

```

Read in the protein details

```{r}
uniprot2details <- readRDS(here('external/uniprot2details_unique.rds'))
```


# Read the peptide data into a QFeatures object

We now start reading in the peptide-level quantification data from MaxQuant into a `QFeatures` object.

## Defining the contaminant proteins

We need to remove contaminant proteins. Below, we parse the MaxQuant contaminants fasta to extract the IDs for the proteins.

```{r}
contaminant_proteins <- get_maxquant_cont_accessions()
```

## Reading in the peptide-level quantification data

We need to read in both the proteinGroups.txt files from MaxQuant to identify the proteins which pass the 1% FDR threshold, and the modificationSpecificPeptides.txt file to get the peptide-level quantification.

The search database for this experiment includes separate entries for the free HALO tag construct (`PER2HALO`/`PER2MNG`) and, for peptides shared between the PER2-HALO fusion and endogenous PER2, MaxQuant does not always assign the UniProt accession (`O15055`) as the leading razor protein. Below we override `Leading.razor.protein` for these two specific cases, so that every peptide from the tag construct is consistently grouped under a single pseudo-protein (`HALO_TAG`) and every peptide from the fusion/endogenous protein is grouped under the PER2 UniProt accession, regardless of what MaxQuant happened to assign.

```{r}
pg_inf <- here('raw/HALO/proteinGroups.txt.gz')
pg_df <- read.delim(pg_inf)


seq2razor <- read.delim(here('raw/HALO/peptides.txt.gz')) %>%
  select(Sequence, Leading.razor.protein)

pep_inf <- here('raw/HALO/modificationSpecificPeptides.txt.gz')
pep_df <- read.delim(pep_inf)

# Merge with the leading razor protein column
pep_df <- merge(pep_df, seq2razor, by='Sequence')

per2_lines <- which(grepl('O15055', pep_df$Proteins))
halo_lines <- which(pep_df$Proteins=='PER2HALO;PER2MNG')

pep_df$Leading.razor.protein[halo_lines] <- 'HALO_TAG'
pep_df$Leading.razor.protein[per2_lines] <- 'O15055'

# Merge with the protein details and Q-values
pep_df <- merge(pep_df, uniprot2details,
                by.x='Leading.razor.protein', by.y='UniprotID', all.x=TRUE) %>%
  merge(pg_df[,c('Majority.protein.IDs', 'Q.value')], 'Leading.razor.protein',
        by.y='Majority.protein.IDs', all.x=TRUE)

```


## Extracting the experiment to sample mapping

MaxQuant refers to each raw file by an internal single-letter "experiment" label rather than a readable sample name. We recover the mapping between the two from the search parameter file (`mqpar.xml`), which lists the raw file paths (from which we parse the intended sample name) alongside the experiment labels in the same order. From the sample name we derive `condition` (HA = HALO-only control, P2H = PER2-HALO) and `replicate`.

```{r}


mqpar_inf <- here('raw/HALO/mqpar.xml')
mqpar <- xml2::read_xml(mqpar_inf)

# Extract the file paths and experiment names
filepaths <- strsplit(
  xml_text(xml_find_all(mqpar, ".//filePaths")), split='F:\\\\')[[1]]
filepaths <- filepaths[filepaths!='']

# Extract sample names from file paths
sample <- gsub(
  '(2565608863_Andrei_PER2_121225\\\\2565608863_Andrei_|.raw)', '', filepaths)

experiments <- strsplit(
  xml_text(xml_find_all(mqpar, ".//experiments")), split='')[[1]]

# Checking that the lengths match
length(sample)==length(experiments)

# Create the coldata dataframe
coldata <- data.frame('sample'=sample, 'experiments'=experiments) %>%
  separate(sample, into=c('sample', NA)) %>%
  distinct() %>%
  mutate(sample=sub('(HA|P2H)', '\\1_', sample)) %>%
  separate(sample, into=c('condition', 'replicate'), sep='_', remove=FALSE) %>%
  mutate(quantCols=sample) %>%
  tibble::column_to_rownames('sample')

# named list to map from experimental name to sample name
exp2sample <- coldata$quantCols
names(exp2sample) <- coldata$experiments

coldata %>%
  knitr::kable()


```

## Creating the QFeatures object

Now we can finally create the QFeatures object. The abundance columns (32-37 in the merged data frame) correspond to the per-experiment `Intensity` columns MaxQuant reports; we rename them from the internal experiment label to the readable sample name using the `exp2sample` mapping derived above, and reorder the sample metadata (`coldata`) to match.

```{r}

# Define the columns corresponding to the abundance values
abundance_cols <- 32:37

# Rename the abundance columns to the sample names
colnames(pep_df)[abundance_cols] <- gsub(
  'Intensity.', '', colnames(pep_df)[abundance_cols])

colnames(pep_df)[abundance_cols] <- sapply(
  colnames(pep_df)[abundance_cols], function(x) exp2sample[[x]])

# Reorder the coldata to match the order of the abundance columns
coldata <- coldata[colnames(pep_df)[abundance_cols],]

# Create the QFeatures object
lfq_qf <- readQFeatures(assayData = pep_df,
                        colData = coldata,
                        quantCols = abundance_cols,
                        name = 'peptides_raw')

# Set the colData for the peptides_raw experiment
colData(lfq_qf[['peptides_raw']]) <- colData(lfq_qf)
```

MaxQuant has reported `r sum(assay(lfq_qf[['peptides_raw']])==0, na.rm=TRUE)` quantification values as zero. These should be replaced with NA, since mass spectrometry is not capable of asserting that the protein had zero abundance in the sample.

```{r}
lfq_qf[['peptides_raw']] <- QFeatures::zeroIsNA(lfq_qf[['peptides_raw']])

```




# Filter the peptide-level quantification data

We will now filter the peptide-level quantification data to remove contaminants and proteins which do not pass the 1% FDR threshold.


```{r}

lfq_qf[['peptides_filtered']] <- biomasslmb::filter_features_mq_dda(
  lfq_qf[['peptides_raw']],
  contaminant_proteins=contaminant_proteins,
  filter_contaminant=TRUE,
  filter_associated_contaminant=TRUE,
  unique_master=FALSE)

lfq_qf <- filterFeatures(lfq_qf, i='peptides_filtered', ~ Q.value <= 0.01)

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_filtered']]),
                           "Leading.razor.protein",
                           "Proteins with Q-Value > 0.01 removed")
```

# Normalise the peptide-level quantification data

For the normalisation and summarisation, we need our quantification
values to be on a log-scale, so we log-transform below

```{r}
lfq_qf[['peptides_filtered']] <- QFeatures::logTransform(
  lfq_qf[['peptides_filtered']], base=2)
```

Next, we plot the peptide intensities.

```{r, warning=FALSE, fig.show='hold', fig.show="hold", out.width="50%"}

# Plot the peptide-level quantification distributions per sample
plot_quant(lfq_qf[['peptides_filtered']], log2transform=FALSE, method='density') +
  scale_colour_manual(values=rep(get_cat_palette(4), each=3)) +
  aes(linetype=sample) +
  scale_linetype_manual(values=rep(1:3, times=4)) +
  xlab('Peptide abundance (log2)')
```

## Normalise

Since the same amount of sample was injected for each sample, we could
normalise to remove differences in the overall distribution of
abundances. This is performed below using median normalisation.

```{r}
lfq_qf[['peptides_norm']] <- QFeatures::normalize(lfq_qf[['peptides_filtered']],
                                                  method='diff.median')
```

Below, we can check this normalisation step has worked as intended.

```{r, warning=FALSE, fig.show='hold', fig.show="hold", out.width="50%"}

# Plot the peptide-level quantification distributions per sample
plot_quant(lfq_qf[['peptides_norm']], log2transform=FALSE, method='density') +
  scale_colour_manual(values=rep(get_cat_palette(4), each=3)) +
  aes(linetype=sample) +
  scale_linetype_manual(values=rep(1:3, times=4)) +
  xlab('Peptide abundance (log2)')
```

## Normalise to histone proteins

Median normalisation is suitable when we expect the overall distribution of protein abundances to be similar between the conditions. However, we don't expect the overall distribution of protein abundances to neccessarily be similar between the conditions here, since one is a free HALO tag and the other is PER2-HALO. We should also consider normalising to a set of proteins which we expect to be constant across the samples. Histone proteins are often used for this purpose since they are generally stable in abundance. Below, we identify the histone proteins in our dataset and explore their abundance post normalisation. They are more abundant in the HALO condition, suggesting that global median normalisation may be more inappropriate.

```{r}
hist_df <- uniprot2details %>%
  filter(grepl('^Histone H', Protein.names))

hist_df %>%
  select('Uniprot'=UniprotID, 'Gene Name'=Gene.Names.First, 'Protein names'=Protein.names) %>%
  knitr::kable()

hist <- hist_df %>%
  pull(UniprotID)

# Plot the peptide-level quantification distributions per sample
plot_quant(lfq_qf[['peptides_norm']][
  rowData(lfq_qf[['peptides_norm']])$Leading.razor.protein %in% hist,],
  log2transform=FALSE, method='density') +
  scale_colour_manual(values=rep(get_cat_palette(4), each=3)) +
  aes(linetype=sample) +
  scale_linetype_manual(values=rep(1:3, times=4)) +
  xlab('Peptide abundance (log2)')

```


Below, we normalise to the histone proteins and replot the peptide-level abundance distributions. This is the normalised peptide-level assay (`peptides_norm_hist`) that is carried forward into notebook 2 for protein-level summarisation.

```{r}
hist_medians <- get_medians(lfq_qf[['peptides_filtered']][
  rowData(lfq_qf[['peptides_filtered']])$Leading.razor.protein %in% hist,])

lfq_qf[['peptides_norm_hist']] <- center_normalise_to_ref(
  lfq_qf[['peptides_filtered']], medians = hist_medians, on_log_scale = TRUE)

# Plot the peptide-level quantification distributions per sample
plot_quant(lfq_qf[['peptides_norm_hist']], log2transform=FALSE, method='density') +
  scale_colour_manual(values=rep(get_cat_palette(4), each=3)) +
  aes(linetype=sample) +
  scale_linetype_manual(values=rep(1:3, times=4)) +
  xlab('Peptide abundance (log2)')
```

# Removing peptides with too many missing values

For the estimation of protein-level abundances in the next notebook, we will use `limpa`. We do not need to exclude
peptides with any missing values for this, since `limpa` explicitly models missingness via a detection probability
curve rather than requiring complete data. Below we nonetheless inspect how much of the peptide set would be lost
under a conventional missing-value filter, as a diagnostic; the resulting filtered assay is used only for the
peptide/protein-count QC plots at the end of this notebook and is not itself passed on to notebook 2.

Below, we inspect the most common patterns for the missing values.

The missingness is sometimes consigned to all samples from a condition(s). It looks like the missingness is a mixture of missing at random (MAR) and missing not at random (MNAR) and is likely partly due to falling below detection limits.



```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE, out.width='70%'}
plot_missing_upset(lfq_qf, i='peptides_norm_hist')
```

Here, we'll set a limit of maximum 4/6 missing values per peptide.

```{r}
table(nNA(lfq_qf[['peptides_norm_hist']])$nNArows$nNA)

lfq_qf[['peptides_filtered_missing']] <- QFeatures::filterNA(
  lfq_qf[['peptides_norm_hist']], 4/6)

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_filtered_missing']]),
                           'Leading.razor.protein',
                           "Removing peptides with more than 4/6 missing values")



```



# Inspecting the number of peptides and proteins through the processing steps

Now that we have performed all the filtering steps, it's helpful to visualise how many peptides/proteins were retained at each level of the processing.

## Samples per peptide

We'll start by inspecting the number of peptides in each experiment. We set the row variables to be `Sequence` and `Modifications` so that we count the number of unique peptides.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, out.width='75%', warning=FALSE, message=FALSE}


rename_cols <- c('All peptides' = 'peptides_raw' ,
                 'Quantified, contaminants removed' = 'peptides_filtered',
                 'Max 4/6 missing values' = 'peptides_filtered_missing')

rowvars <- c('Sequence', 'Modifications')
rowDataNames(lfq_qf)[['peptides_raw']]

samples_present <- get_samples_present(
  lfq_qf[,,unname(rename_cols)], rowvars, rename_cols)

plot_samples_present(samples_present, rowvars, breaks=seq(2,10,2)) + ylab('PSM')


```

### Samples per Protein

Next, we'll use the same functions to inspect the number of proteins at each level of processing. We need to set the row variables to be just the `Leading.razor.protein` column.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, out.width='75%', warning=FALSE, message=FALSE}


rowvars_prot <- c('Leading.razor.protein')

samples_present <- get_samples_present(lfq_qf, rowvars_prot, rename_cols)
plot_samples_present(samples_present, rowvars_prot, breaks=seq(2,10,2))
```


# Save outputs

Save the full `QFeatures` object (all processing stages) for use in notebook 2.

```{r}
output_dir <- here('notebooks/output')

saveRDS(lfq_qf, here(output_dir, 'HALO_lfq_qf.rds'))
```


# Session info

```{r}
sessionInfo()
```
