---
title: "PER2-HALO: Data exploration and statistical testing"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: pdf_document
---

Here, we explore the protein-level quantitative data and perform statistical testing.

# Setup

Read in the required packages

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(QFeatures)
library(biomasslmb)
library(here)
library(limpa)
library(limma)
library(ggplot2)
library(dplyr)
library(tidyr)


```

Define directory to save plots

```{r}
plots_dir <- here('results/plots')
if(!dir.exists(plots_dir)){
  dir.create(plots_dir, recursive=TRUE)
}
```

Read in the quantification data.

```{r}
lfq_qf <- readRDS(here('notebooks/output/HALO_lfq_qf.rds'))
uniprot2details_unique <- readRDS(here('external/uniprot2details_unique.rds'))
```

Define proteins to highlight

```{r, message=FALSE, warning=FALSE}
mtor_proteins_to_highlight <- readxl::read_excel(
  here('raw/HALO/PER2 IP - List for volcano plot.xlsx'), col_names = FALSE)

clock_proteins <- c('PER2', 'PER1', 'CRY1', 'CRY2', 'CSNK1D',
                    'CSNK1E', 'CLOCK', 'ARNTL', 'ARNTL1', 'ARNTL2')


```

# Summarising to protein-level using `limpa`

The first step is to perform protein quantification using `limpa`. We start by creating an `EList` object from the peptide level data, then fitting the Detection Probability Curve (DPC) model and performing protein quantification. The DPC is used to model the relationship between peptide detectability and abundance, allowing for more accurate protein quantification in the presence of missing values.

```{r, out.width='50%'}

y.peptide = new("EList", list(E=assay(lfq_qf[['peptides_norm_hist']]),
                              genes=rowData(lfq_qf[['peptides_norm_hist']])))

# Filter to proteins with at least 2 peptides
y.peptide = filterCompoundProteins(y.peptide,
                                   protein.group = "Leading.razor.protein",
                                   min.n.peptides=2)
dpcfit <- dpc(y.peptide)
plotDPC(dpcfit)


y.protein <- dpcQuant(y.peptide, protein.id="Leading.razor.protein", dpc=dpcfit)

```

Next, we want to add the protein-level quantification results from `limpa` back into our `QFeatures` object for further analysis. For this, we need to create a 1) `summarizedExperiment` object, 2) Add it to our `QFeatures` object and 3) add assay links to the ppeptide-level data.

```{r, warning=FALSE, message=FALSE}
# Create a temporary protein-level summarizedExperiment so we can extract the
# rowData for our limpa summarizedExperiment
lfq_qf <- aggregateFeatures(lfq_qf, 
                            i = "peptides_norm_hist", 
                            fcol = "Leading.razor.protein",
                            name = "protein_sum",
                            fun = colSums, na.rm=TRUE)

lfq_qf <- QFeatures::zeroIsNA(lfq_qf, i='protein_sum')

# Create the limpa protein-level summarizedExperiment
limpa_protein <- SummarizedExperiment(
  assays=list(y.protein$E),
  rowData = rowData(lfq_qf[['protein_sum']])[rownames(y.protein),],
  colData = colData(lfq_qf))

# Add the limpa protein-level assay to the QFeatures object
lfq_qf <- addAssay(x = lfq_qf, 
                   y = limpa_protein, 
                   name = "protein_limpa")

# Add assay link from peptide-level data to protein-level data
lfq_qf <- addAssayLink(object = lfq_qf, 
                      from = "peptides_norm_hist", 
                      to = "protein_limpa",
                      varFrom='Leading.razor.protein',
                      varTo='Leading.razor.protein')

```

Below, we extract the standard errors from the `limpa` results for later use in plotting.

```{r}
# Extract standard errors and expression values in long format
stderrors_long <- y.protein$other$standard.error %>%
  data.frame() %>%
  tibble::rownames_to_column('UniprotID') %>%
  pivot_longer(cols=-'UniprotID', values_to='std_error', names_to='sample') %>%
  mutate(sample=remove_x(sample))

# Extract expression values in long format
exp_long <- y.protein$E %>%
  data.frame() %>%
  tibble::rownames_to_column('UniprotID') %>%
  pivot_longer(cols=-'UniprotID', values_to='exp', names_to='sample') %>%
  mutate(sample=remove_x(sample))

# Create long form data for plotting, including a column stating whether a protein
# quantification was 'imputed', which here means inferred without quantification
# of any peptides for the protein in that sample
long_protein <- biomasslmb::create_long_form_imputed_data(
  lfq_qf, i_imputed='protein_limpa', i_unimputed = 'protein_sum',
  row_variables = c('Leading.razor.protein', 'Gene.Names.First'),
  column_variables = c('condition', 'replicate'))

# Merge in the standard errors
long_protein_inc_std_error <- merge(
  long_protein,
  stderrors_long, by.x=c('primary', 'rowname'),
  by.y=c('sample', 'UniprotID')) 

# Save long format protein quantification with standard errors
long_protein_inc_std_error %>%
  mutate(any_pep_quant=!is.imputed) %>%
  select(Leading.razor.protein, Gene.Names.First, condition, replicate, abundance=value, std_error, any_pep_quant) %>%
  write.table(file = here('results/HALO_long_format_protein.tsv'),
              quote=FALSE, row.names=FALSE, sep='\t')

# We can now delete the temp SE
lfq_qf[['protein_sum']] <- NULL

```


Save the protein quantification in wide format

```{r}

protein_wide_format <- assay(lfq_qf[['protein_limpa']]) 

protein_wide_format <- protein_wide_format %>%
  data.frame() %>%
  tibble::rownames_to_column('UniprotID')


write.table(protein_wide_format,
            file = here('results/HALO_wide_format_protein.tsv'),
            quote=FALSE, row.names=FALSE, sep='\t')
```




# Data exploration

Before we proceed to statistical testing, we first explore the data to ensure that the experimental conditions have had the expected effect on the proteome.

## Correlation between samples


We start with a plot of the correlation matrix, with the samples ordered by a hierachical clustering of the distance between them. The samples clearly separate by condition

```{r, fig.height=6, fig.width=6, out.width="50%"}

corrplot::corrplot(cor(assay(lfq_qf[['protein_limpa']]),use='complete',
                       method='spearman'),
           col = corrplot::COL2('RdBu'),
           method = 'pie',
           type = 'upper',
           order='hclust',
           tl.cex=1,
           cl.cex=1, tl.col='grey20', is.corr=FALSE)

```

## Principle component analysis

Below, we use PCA to examine the major sources of variance.

```{r, fig.show='hold', out.width="50%"}

plot_pca(lfq_qf, 'protein_limpa',
         scale=TRUE, center=TRUE) + 
  aes(fill=condition) +
  #scale_shape_manual(name='', values=c(21,22)) +
  theme_biomasslmb(base_size=10)



```
PCA gives the same overall view of the samples, with the separation between conditions accounting for ~70% of the variance



# Statistical testing

Having established that the experimental conditions appear to have had the expected effect, we proceed to statistical testing for PER2-HALO (P2H) vs HALO (HA). We use the `limma` package for this, applying the empirical Bayes moderation to the standard errors to improve statistical power.

```{r, out.width='50%'}

# Design matrix
condition <- factor(lfq_qf$condition, levels=c("HA", "P2H"))

design <- model.matrix(~ condition)

fit <- dpcDE(y.protein, design)


## Update the model using the limma eBayes algorithm
final_model <- eBayes(fit = fit, 
                    trend = TRUE,
                    robust = TRUE)

plotSA(final_model)

## Format results
limpa_results <- topTable(
  fit = final_model,   
  coef = 2, confint = TRUE,
  adjust.method = "BH",    # Method for multiple hypothesis testing
  number = Inf)

# Check distribution of p-values
limpa_results %>%
  as_tibble() %>%
  ggplot(aes(x = P.Value)) + 
  geom_histogram() +
  theme_biomasslmb()
```



How many significantly changing proteins at 5% FDR?

```{r}
table(ifelse(limpa_results$logFC>0, 'Inc.', 'Dec.'),
      sig_5_perc_FDR=limpa_results$adj.P.Val<0.05)
```

# Visualisation of results

## Volcano plot highlighting clock proteins and mTOR pathway proteins

```{r, warning=FALSE, message=FALSE}


highlight_clock <- limpa_results %>%
  filter(Gene.Names.First %in% clock_proteins) %>%
  mutate(significant=adj.P.Val<0.05 & logFC>0)
  
highlight_mtor <- limpa_results %>%
    filter(Gene.Names.First %in% union(mtor_proteins_to_highlight$`...1`,
                                     mtor_proteins_to_highlight$`...2`) |
        Leading.razor.protein %in% mtor_proteins_to_highlight$`...3`) %>%
  mutate(significant=adj.P.Val<0.05 & logFC>0)

p <- limpa_results %>%
  filter(is.finite(P.Value)) %>%
  mutate(significant=adj.P.Val<0.05 & logFC>0) %>%
  bind_rows(.id='enrichment_method') %>%
  ggplot(aes(logFC, -log10(P.Value), alpha=significant, size=significant)) +
  geom_point(stroke=0.25, shape=20, aes(colour=significant)) +
  geom_point(size=2, pch=21, fill=get_cat_palette(1), data=highlight_clock) +
  geom_point(size=2, pch=21, fill=get_cat_palette(2)[2], data=highlight_mtor) +
  ggrepel::geom_text_repel(
    data=bind_rows(highlight_clock, highlight_mtor),
    aes(label=ifelse(Leading.razor.protein=='HALO_TAG', 'HALO_TAG',
                     gsub(' .*', '', Gene.Names.y))),
    max.overlaps = 50, size=4, min.segment.length =unit(0, 'mm'),
    colour='grey20', alpha=1) +
  theme_biomasslmb() +
  scale_colour_manual(values=c('grey80', 'grey60'),guide=FALSE) +
  theme(strip.background = element_blank()) +
  scale_size_manual(values=c(1, 2), guide='none') +
  scale_alpha_manual(values=c(0.25, 1), guide='none') +
  labs(x=expression(Log[2]*FC~"(PER2-HALO/HALO)"),
         y=expression(-log[10] (p-value))) +
  scale_x_continuous(breaks=seq(-10,10,2))


print(p)

ggsave(here(plots_dir, 'limpa_volcano_per2h_vs_halo.png'))
ggsave(here(plots_dir, 'limpa_volcano_per2h_vs_halo.pdf'))

limpa_results %>%
  select(Leading.razor.protein, Entry.Name, Protein.names, Gene.Names=Gene.Names.y,
         Gene.Names.First, NPeptides, logFC, CI.L, CI.R, AveExpr,
         P.Value, adj.P.Val) %>%
  write.table(here('results/HALO_limpa_results.tsv'), sep='\t',
              row.names=FALSE, quote=FALSE)
```

## Protein abundances for clock proteins and mTOR pathway proteins


```{r, warning=FALSE, message=FALSE}
to_plot <- limpa_results %>%
  arrange(-sign(logFC)*-log10(P.Value)) %>%
  mutate(rank=1:n()) %>%
  filter((Gene.Names.First %in% clock_proteins)|
           Gene.Names.First %in% union(mtor_proteins_to_highlight$`...1`,
                                              mtor_proteins_to_highlight$`...2`)|
                    (Leading.razor.protein %in% mtor_proteins_to_highlight$`...3`))
           
protein_order <- c('PER2', setdiff(sort(clock_proteins), 'PER2'))
protein_order <- c('PER2', setdiff(to_plot$Gene.Names.First,'PER2'))
                   
p <- to_plot %>%
  mutate(Gene.Names.First=factor(Gene.Names.First,levels=protein_order)) %>%
  mutate(type=case_when(
    Gene.Names.First %in% clock_proteins~'Clock protein',
    Gene.Names.First %in% c('MTOR', 'MLST8', 'RICTOR') ~ 'mTOR complex',
    .default = 'mTOR regulators')) %>%
  mutate(type=factor(
    type, levels=c('Clock protein', 'mTOR complex', 'mTOR regulators'))) %>%
  ggplot(aes(Gene.Names.First, logFC)) +
  geom_hline(yintercept = 0, linetype='dashed', colour='grey20') +
  geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width=0.2, colour='grey40') +
  geom_point(colour='white', size=4) +
  geom_point(aes(alpha=adj.P.Val<0.05, colour=type), size=3) +
  theme_biomasslmb(aspect_square = FALSE) +
  labs(x='',
       y=expression(Log[2]*FC~"(PER2-HALO/HALO)")) +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1)) +
  
  scale_colour_manual(values=c('grey80', get_cat_palette(5)[5]),
                      name='Significant\n(FDR<5%)') +
  facet_grid(.~type, scales='free_x', space='free_x') +
  theme(strip.background = element_blank()) +
  scale_colour_manual(values=rep(get_cat_palette(2),
                                 times=c(1,2)), guide='none') +
  scale_alpha_manual(values=c(0.25, 1), guide='none') +
  theme(panel.spacing.x = unit(5, 'mm'))
print(p)

ggsave(here(plots_dir, 'limpa_clock_mtor_per2h_vs_halo.png'))
ggsave(here(plots_dir, 'limpa_clock_mtor_per2h_vs_halo.pdf'))
```

Below, we inspect the abundance values for individual samples for these proteins. For some like CRY1/CRY2, we haven't quantified any peptides in any HALO controls samples, so the protein abundance is inferred from the `limpa`` model in the absence of quantification. The fold-changes in these cases are obviously less reliable, but given all the PER2-HALO samples are well quantified, it does suggest a strong enrichment. Aside from MIOS, all the mTOR pathway proteins highlighted are quantified in some peptides for at least 2/3 HALO ciontrol replicates, which gives us reasonable confidence in the fold-changes observed.

```{r, fig.width=12, fig.height=6}

long_protein_inc_std_error %>%
  filter((Gene.Names.First %in% clock_proteins)|
           Gene.Names.First %in% union(mtor_proteins_to_highlight$`...1`,
                                              mtor_proteins_to_highlight$`...2`)|
                    (Leading.razor.protein %in% mtor_proteins_to_highlight$`...3`)) %>%
  mutate(Gene.Names.First=ifelse(
    Leading.razor.protein=='HALO_TAG', 'HALO_TAG-HALO', Gene.Names.First)) %>%
  ggplot(aes(Gene.Names.First, value, fill=condition,
             group=interaction(replicate, condition), shape=is.imputed)) +
  geom_errorbar(aes(ymin=value - std_error, ymax=value + std_error),
                position=position_dodge(width = 0.5), width=0.2, colour='grey') +
  geom_point(position=position_dodge(width = 0.5), size=2, colour='grey20') +
  theme_biomasslmb(border=FALSE) +
  theme(strip.background=element_blank(), plot.title=element_text(hjust=0.5)) +
  scale_shape_manual(values=c(21, 25),
                     labels=c('\nQuantified\npeptides\n', '\nInferred without\nquantification\n'),
                     name='') +
  scale_fill_manual(values=get_cat_palette(4)[3:4], name='') +
  labs(x='', y='Protein abundance (log2)') +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1),
        aspect.ratio = 1/4) +
  guides(fill=guide_legend(override.aes = list(size=4, pch=21)))
```


```{r}
sessionInfo()
```
