---
title: "PER2-PROTAC: Data exploration and statistical testing"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: pdf_document
---

Here, we explore the protein-level quantitative data and perform statistical testing to identify phosphorylation changes upon PROTAC treatment.

# Setup

Load required packages

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(QFeatures)
library(biomasslmb)
library(here)
library(limma)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(ggbeeswarm)

```


Define directory for plots

```{r}
plots_dir <- here('results/plots')
if(!dir.exists(plots_dir)){
  dir.create(plots_dir, recursive = TRUE)
}
```

Read in the data

```{r}

output_dir <- here('notebooks/output')

qfs <- readRDS(here(output_dir, 'PROTAC_qfs.rds'))

long_format_protein <- readRDS(
  here(output_dir, 'PROTAC_long_format_protein.rds'))

long_long_format_phosphopeptide_norm_to_protein <- readRDS(
  here(output_dir, 'PROTAC_long_long_format_phosphopeptide_norm_to_protein.rds'))


```


# Data exploration

## PCA plots

Below, we make PCA plots for the protein and phosphopeptide data. At the protein level, DMSO and PROTAC separate on PC1 (30% of variance), with two DMSO-2 and PROTAC-3 relatively close together. For phosphopeptides, the conditions separate very well (PC1; ~40% variance explained)

```{r, fig.show='hold', out.width='50%', warning=FALSE, message=FALSE}
for(qf in names(qfs)){

  if(qf=='total'){
    assay_name <- 'protein'
    feature <- 'Protein'
  } else{
    assay_name <- 'phosphopeptide'
    feature <- 'Phosphopeptide'
  }
  
  
  print(plot_pca(
    qfs[[qf]], i=assay_name, colour_by = 'condition', shape='replicate') +
      ggtitle(feature))

}
```

# Statistical testing of protein abundance changes between conditions


## Protein

First, we'll perform statistical testing to compare the conditions (PROTAC vs DMSO) for total protein abundance

```{r, out.width='50%', warning=FALSE, message=FALSE, fig.show='hold'}

## Create design matrix for the linear model
condition <- factor(qfs$total[['protein']]$condition, levels=c('DMSO', 'PROTAC'))
m_design <- model.matrix(~condition)

## Fit linear model using the design matrix and desired contrasts
fit_model <- lmFit(object = assay(qfs$total[['protein']]), design = m_design)

## Update the model using the limma eBayes algorithm
final_model <- eBayes(fit = fit_model, 
                      trend = TRUE,
                      robust = TRUE)

## Add TREAT to final model to perform fold-change thresholding for contrasts
# require |FC| significantly over 1.25x
final_model_treat <- treat(fit_model, fc=1.25, trend=TRUE, robust=TRUE)  

## QC plot to assess if there is a trend between abundance and variance, as expected
plotSA(fit = final_model,
       cex = 0.5,
       xlab = "Average log2 abundance")

## Format results
prot_limma_results <- topTreat(
  fit = final_model_treat,   
  coef = 'conditionPROTAC', 
  adjust.method = "BH",    # Method for multiple hypothesis testing
  number = Inf) %>%        # Print results for all proteins
  tibble::rownames_to_column("Protein") %>%
  merge(data.frame(rowData(qfs$total)[['protein']]),
        by.x='Protein', by.y='row.names', all.x=TRUE) %>%
  arrange(P.Value)

## Check distribution of p-values. Because we are using treat, we don't expect a 
## uniform distribution under the null, but rather a peak near 1.
p <- prot_limma_results %>%
  as_tibble() %>%
  ggplot(aes(x = P.Value)) + 
  geom_histogram() +
  theme_biomasslmb()

print(p)

```


How many proteins have a significant change in abundance

```{r}
table(prot_limma_results$adj.P.Val<0.05,
      ifelse(prot_limma_results$logFC>0, 'Increased', 'Decreased')) %>%
  knitr::kable()

prot_limma_results %>%
  filter(adj.P.Val<0.05) %>%
  select(Protein, Gene.Names.First, logFC, adj.P.Val) %>%
  knitr::kable()
```

OK, so very few proteins have a significant change in abundance!




### Phosphopeptide statistical testing

Next, we perform statistical testing for the phosphopeptide data, comparing PROTAC vs DMSO while accounting for the matched protein abundance measurements.

```{r}
# Create data matrix for modelling
to_model_phospho <- long_long_format_phosphopeptide_norm_to_protein %>%
  filter(type!='phosphopeptide_prot_norm') %>%
  select(prot_phos, condition, replicate, type, abundance) %>%
  pivot_wider(names_from = c(condition, replicate, type), values_from=abundance) %>%
  tibble::column_to_rownames('prot_phos')

# Create design matrix for the linear model
exp_cond_phospho <- data.frame(sample=colnames(to_model_phospho)) %>%
  separate(sample, into=c('condition', 'replicate', 'type'), remove = FALSE)

condition <- factor(exp_cond_phospho$condition, levels=c('DMSO', 'PROTAC'))

type <- factor(exp_cond_phospho$type, levels=c('protein', 'phosphopeptide'))

m_design_phospho <- model.matrix(~ condition * type)

# same ID for phospho + total from the same sample
block <- interaction(condition, exp_cond_phospho$replicate)   

# Estimate the correlation between measurements from the same block
corfit <- duplicateCorrelation(as.matrix(to_model_phospho),
                               m_design_phospho, block=block)

# Fit the linear model with blocking and correlation information
limma_fit_phospho <- lmFit(
  as.matrix(to_model_phospho), m_design_phospho, block=block, correlation=corfit$consensus)

# Update the model using the limma eBayes algorithm
limma_fit_phospho_eb <- eBayes(limma_fit_phospho, trend=TRUE, robust=TRUE)  

# QC plot to assess if there is a trend between abundance and variance, as expected,
# and if the limma 'trend' captures this
plotSA(limma_fit_phospho_eb)

# Format results
phospho_limma_results <- topTable(limma_fit_phospho_eb, coef = 'conditionPROTAC:typephosphopeptide',
  adjust.method = "BH",    # Method for multiple hypothesis testing
  number = Inf) %>%        # Print results for all proteins
  tibble::rownames_to_column("phosphosite") %>%
  merge(rowData(qfs$phospho)[['phosphopeptide']][
    ,c('Entry.Name', 'Protein.names.y', 'Gene.Names', 'Gene.Names.First', 'site_seq')],
        by.x='phosphosite', by.y='row.names', all.x=TRUE) %>%
  data.frame() %>%
  separate(phosphosite, into=c('UniprotID', 'phospho_pos'), sep='_', remove=FALSE)


```



Save results

```{r}
saveRDS(phospho_limma_results, here(output_dir, 'PROTAC_phospho_limma_results.rds'))

```

## How many phosphosites have a significant change in abundance?

Below, we summarise how many phosphosites have a significant change in abundance (FDR<5%)

```{r, out.width='50%'}
table(ifelse(phospho_limma_results$adj.P.Val<0.05, 'Significant', 'Not significant'),
      ifelse(phospho_limma_results$logFC>0, 'Increased', 'Decreased')) %>%
  knitr::kable()


phospho_limma_results %>%
  mutate(type=ifelse(grepl('Y', phospho_pos), 'Tyrosine', 'Serine/Threonine')) %>%
  mutate(sig=adj.P.Val<0.05) %>%
  group_by(sig, type, direction=ifelse(logFC>0, 'Increased', 'Decreased')) %>%
  tally() %>%
  group_by(type, direction) %>%
  mutate(perc=100*n/sum(n)) %>%
  filter(sig) %>%
  ungroup() %>%
  complete(direction, type, fill=list(n=0)) %>%
  ggplot(aes(type, perc, fill=direction)) +
  geom_col(position='dodge', width=0.5) +
  theme_biomasslmb() +
  labs(x='', y='Percentage of significant\nphosphosites (FDR<5%)', fill='Direction of change') +
  scale_fill_manual(values=get_cat_palette(2)) +
  geom_text(aes(label=n), position=position_dodge(width=0.5), vjust=-0.5, size=4) +
  ylim(0, 40) +
  theme(strip.background = element_blank())
```

Below, we make a volcano plot for the statistical test results

```{r, fig.width=6, fig.height=6}
# highlight top 10 significant proteins per genotype
phospho_limma_results %>%
  group_by(logFC>0) %>%
  filter(adj.P.Val<0.05) %>%
  slice_min(P.Value, n=10) -> highlight

phospho_limma_results %>%
  mutate(significant=adj.P.Val<0.05) %>%
  ggplot(aes(logFC, -log10(P.Value))) +
  geom_point(aes(fill=significant,
             alpha=significant, size=significant),
             colour='grey20', stroke=0.25, shape=21) +
  theme_biomasslmb() +
  scale_fill_manual(values=c('grey60', get_cat_palette(1)),
                      name='Significant\n(FDR<5%)') +
  scale_size_manual(values=c(0.5, 1), guide='none') +
  scale_alpha_manual(values=c(0.1, 0.5), guide='none') +
  labs(x='log2 FC (PROTAC vs DMSO)', y='-log10(p-value)') +
  theme(strip.background = element_blank()) +
  geom_text_repel(
    data=highlight,
    aes(label=paste(Gene.Names.First, gsub('.*_', '', phosphosite))),
    max.overlaps = 50, size=3, min.segment.length =0, colour='black', force = 10) +
  ggtitle('Phosphorylation') -> p

print(p)
```

## Plot changes in phosphorylation for AKT1S1

Below, we plot the changes in phosphorylation for AKT1S1, which has multiple significantly changing sites

```{r}

p_values <- phospho_limma_results %>%
  filter(Gene.Names.First=='AKT1S1') %>%
  mutate(ptm_name=phosphosite)

to_plot <- long_long_format_phosphopeptide_norm_to_protein %>%
  filter(Gene.Names.First=='AKT1S1', type=='phosphopeptide_prot_norm')

DMSO_abundances <- to_plot %>%
  filter(condition=='DMSO') %>%
  group_by(prot_phos) %>%
  summarise(abundance=mean(abundance))

to_plot <- to_plot %>%
  merge(DMSO_abundances, by='prot_phos', suffix=c('', '_DMSO')) %>%
  mutate(abundance_diff=abundance - abundance_DMSO) %>%
  group_by(condition, prot_phos) %>%
  mutate(mean_abundance_diff=mean(abundance_diff)) %>%
  arrange(desc(mean_abundance_diff)) %>%
  mutate(ptm_name_facet=gsub('; ', ';\n', ptm_name))

to_plot$ptm_name_facet <- factor(to_plot$ptm_name_facet,
                                 levels=unique(to_plot$ptm_name_facet))

p_values <- phospho_limma_results %>%
  filter(Gene.Names.First=='AKT1S1') 

sig_sites <- p_values %>%
  filter(adj.P.Val<0.1) %>%
  pull(phospho_pos)

p <- to_plot %>%
  filter(ptm_name %in% sig_sites) %>%
  filter(condition!='DMSO') %>%
  ggplot(aes(ptm_name, abundance_diff)) +
  geom_hline(yintercept = 0, colour='grey50', linetype='dashed') +
  geom_point(colour='black', stroke=0.5, shape=21,
             fill=get_cat_palette(2)[2], size=5) +
  stat_summary(fun.min=mean, fun.max=mean, geom='errorbar',
               linewidth=2, width=0.4, colour='grey20') +
  theme_biomasslmb(aspect_square = FALSE) +
  labs(x='', y=expression(Log[2]*FC~"(PROTAC/DMSO)"))  +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
  facet_grid(.~ptm_name_facet, scales='free_x') +
  theme(strip.background = element_blank(),
        panel.spacing = unit(0, 'mm'),
        axis.text.x = element_blank(),
        aspect.ratio=4,
        strip.text.x=element_text(vjust=1, hjust=0),
        plot.title = element_text(hjust=0.5)) +
  ggtitle('AKT1S1')

p1 <- p + theme(strip.text.x=element_text(vjust=0.5, hjust=0.5, size=15))
print(p1)
ggsave(plot=p1, here('results/plots/PROTAC_AKT1S1_phosphosite_changes_selected.png'))
ggsave(plot=p1, here('results/plots/PROTAC_AKT1S1_phosphosite_changes_selected.pdf'))


p2 <- p + filter(to_plot, condition!='DMSO')
print(p2)
ggsave(plot=p2, here('results/plots/PROTAC_AKT1S1_phosphosite_changes.png'))
ggsave(plot=p2, here('results/plots/PROTAC_AKT1S1_phosphosite_changes.pdf'))

```

# Save results

```{r}
phospho_limma_results %>%
  write.table(here('results/PROTAC_phospho_limma_results.tsv'),
              sep='\t', row.names=FALSE, quote=FALSE)
```


Finally, session info:

```{r}
sessionInfo()
```
